Keep your Laziness in Check

### 2018-02-06 paper/main.tex

\hzh{Rewrite counter: 4}
\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}
\leo{I want to keep a counter on how many times we'll rewrite this intro :)}

@@

Lazy evaluation brings tremendous power to functional programming
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks rely on it\cn \leo{This
point might need more exposition about frp}. Haskell is the most
popular programming language that allows programmers to enjoy the
power of lazy evaluation by default. Let us consider a few examples
that demonstrate the power of laziness, and remind us of the
intricacies of programming with laziness in Haskell.

@@

Purely functional queue is a very popular data structure in
Haskell. Programmers typically implement such queues with two
lists---a front list holding the queue elements in FIFO order, and a
back list holding the other elments in reversed order. When the front
list is empty, a dequeue operation requires a reversal of the back
list to refill the front list. These unlucky dequeue operations take
$O(n)$ time to finish, and are much slower than their imperative
counterparts.

However, Okasaki designed a very clever purely functional queue using
lazy lists that both incrementally performs the reversal of the back
list, and incrementally appends the reversal of the back list onto the
front list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
rotate []    back acc = (head back)  : acc
rotate front back acc = (head front) : rest
  where rest = rotate (tail front) (tail back) ((head back) : acc)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, the
function \lk{rotate} is much lazier than the naive reverse and append
operation. The function \lk{reverse} has to evaluate the entire
structure of the back list, which is not incremental. On the other
hand, the function \lk{rotate} produces a list immediately, and packs
the rest of the computation into the lazy list \lk{rest}.

>>>>

%\hzh{Rewrite the intro using different running examples, foldl vs foldr on stream manipulations?}
%\hzh{We can write a hand-crafted spec for take, can we show that we can create this spec by composing the spec for foldr (but not using the one with foldl)}
%\hzh{implement take in 3 ways: foldl, foldr and explicitly as the opening example, and then ask if they're equivalent? (they're not!)}
\hzh{Okasaki has 2 functions that appends a list to the reversal of another list, one of which is too strict and the other one is just the right amount of lazy (thus incremental)}
\hzh{The amortized one is hard to specify exactly, but we can introduce approximate specification and specify it approximately}
\hzh{we need to carefully avoid saying this is a tool for specifying the time complexity of functions}
\hzh{Implementing functional programs that make clever use of laziness is very easy to get wrong, and we provide a solution to make sure the programmer got it right. We almost have no way of testing it right now before StrictCheck.}
\hzh{Email Ryan Trinkle and Dan Winograd-Cort about FRP and strictness, and try to use their examples as another problem we can solve in the intro}
\hzh{Maps will mess up memoization if you're overly strict, use knapsack as an example}

@@

Lazy evaluation gives tremendous power to functional programmers
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks crucially rely on
it\cn \leo{This point might need more exposition about frp}. Haskell
is the most popular programming language that allows programmers to
enjoy the power of lazy evaluation by default. However, that power
does not come without a price: reasoning about lazy programs is
notoriously hard. Let us consider two examples that demonstrate the
power of laziness, as well as the intricacies of programming with it.

@@

\leo{The singular feels weird here. Is it just me?}
Purely functional queue is a very popular data structure in
Haskell. Programmers typically implement such queues with two
lists---a front list holding the queue elements in FIFO order, and a
back list holding the other elments in reversed order. When the front
list is empty, a dequeue operation requires a reversal of the back
list to refill the front list. These specific dequeue operations take
$O(n)$ time to finish, and are much slower than their imperative
counterparts. This occasional performance degradation is compounded by
persistence (where the $O(n)$ operation could be repeated many times),
and can be unacceptable in real-time systems.

However, Okasaki cleverly improved upon this design using lazy lists
that both incrementally performs the reversal of the back list, and
incrementally appends the reversal of the back list onto the front
list ~\cite{Okasaki95simpleand}. Okasaki's queue relies on the
following \lk{rotate} function to incrementally reverse and append the
back list.
\begin{inlinecode}
rotate []    back acc = (head back)  : acc
rotate front back acc = (head front) : rest
  where rest = rotate (tail front) (tail back) ((head back) : acc)
\end{inlinecode}
If we computed the expression \lk{rotate front back []}, its value
will be equivalent to \lk{front ++ (reverse back)}. However, the
function \lk{rotate} is much lazier than the naive reverse and append
operation. The function \lk{reverse} has to evaluate the entire
structure of the back list, which is not incremental. On the other
hand, the function \lk{rotate} produces a list immediately, and packs
the rest of the computation into the lazy list \lk{rest}.
